Formal Construction of Microcontroller Instruction Set Model Using B

### 2010-04-05 papers/Article2010/source/paper.tex
Initial

\documentclass{llncs}

\usepackage{graphicx,array}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}

@@

\title{Formal Verification of Real Models from Platforms Using The B-method }

\author{B. Dantas \inst{1}, David D\'{e}harbe\inst{1}, Val\'{e}rio Medeiros Jr\inst{1} }

\institute{Federal University of Rio Grande do Norte, Natal RN 59078-970, Brazil}

@@

\begin{abstract}
This paper describes an approach to model the microcontroller platforms. More specifically, it shows details about
the Z80 model. The model is created using the B method; it applies math and logic concepts to describe the
characteristics from platforms. Therefore, this model can be used in platforms projects to document, build
simulators, verify properties about the model and verify the software in assembly level. The verification in
assembly is the feature more important, because this allow the developing software faithful to the algorithm
language. Finally, this paper show some relevant techniques used in models and to describe several details.
\end{abstract}
%
%\keywords{Fomal Methods, Microprocessor Verification, Hardware}

\section{Introduction}

The B method allows the construction of safety computacional systems by proofs verification and generation
that guarantees the compatibility among modelling steps and that the system desirable state allways be
true.Therefore, a initial abstract model of the system is defined and refined until the implementation
level. B still allows the code generation from implementation, but this fase isn't verified.

In\cite{Dantas_Report08} is presented a approach to complement the B method until the codify level:
extension of the B method until the assembly level. In this approach, the system construct beggining from
the abstract level and it is formally refined until the target plataform assembly. Therefore, the first
step of this approach is the target plataform modelling, the microprocessor, in B. Using the
responsability division mechanism provide by B basic modules was developted to be used as elementary
library to constructing generic microprocessor, in which bit data type, bit arrays and others are
specified together with its operations\cite{Medeiros}. The resulting processor B machine  will be used as
library in which its operations are used in the modelling of the systems. The use of this library occurs
at the B implementation level of the system.

The processor modelling can be used in plataforms project documentation, simulators construction,
consistency verification and software verification at assembly level. The verification at assembly level
is more interesting utility because allows the developtment of trustworth software by construction, that
is extremaly desirable in embedded critical software.

In this paper we introduced the modelling of Z80 processor\cite{Z80}, since elementary library until
assembly instructions. The paper is structured as follows. Section\ref{B Notation} provides a general introduction
to the B method.Section\ref{General definitions for hardware models} presents the elementary library of
microprocessors basic concepts. Section\ref{Description of the B model from Z80} presents the Z80 B
modelling. Section\ref{} TERMINAR

% Explica o contexto geral do artigo
%  Porque modelar os microcontroladores. \footnote{The interested reader in details from full specification or in
% other study case is invited to visit the repository from authors at \url{http://b2asm.googlecode.com}.
%    Porque podemos verificar várias propriedades em tempo de projeto, documentar, simulacao através das ferramentas de suporte do método B...
%    Porque permite verificar software em nível de assemblagem.[citar trabalho de BHDL]
%    Porque utilizamos o método B puro  - [citar proposta do ALEMAO do ProB]
%  Contextualizacao com grande desafio da verificao formal em nivel de assemblagem[citar grande challenge] 
%  Conclusão da introducao.
%  Organizacao do paper
%  Basear-se no Relatorio técnico e no ERMAC
%
% - Citar o Atelierb

[I need to Review this below text and to answer some questions]
%  What the objectives from this models ?
%  The objectives from this models are: supply a documentation that help the
%  directly in the development and guarantees a series of properties. This
%  properties are about hardware and software that run in this platform. The
%  properties verification about software allow to verify software in the
%  assembly level and It`s provide a high level of security, cause the mains
%  levels of representation of software are verified, according to
%  \cite{DANTAS08_SBMF}.

% Why don´t we use a language based in events, how Event-B or CSP? 
%  Because, we want to prover importants properties about the software and 	
%  the B-method is more appropriate and don´t lost the capacity of verification
%  about hardware level.

% What are the advantages to use the B method ? The biggest advantage is the
% formal verification on assembly level.

% Once established the model of the platform, so I could check something on VHDL?
% Or generate VHDL?

\section{B Notation}

[One page]

\subsubsection{Desing components B }
[half page]
[Explicar como as coias são modeladas]
% Use the tech report

\section{General definitions for hardware models}

We think that We find it convenient to provide a set of basic definitions to model hardware concepts. These
definitions are grouped in a library that we named $\mathit{HARDWARE\_LIBRARY}$ and arrange through four modules:
$\mathit{BIT\_DEFINITION}$, $\mathit{BIT\_VECTOR\_DEFINITION}$, $\mathit{BYTE\_DEFINITION}$ and
$\mathit{BIT\_VECTOR\_ARITHMETICS}$.

[[Explain better the figure]]

The dependency graph between these modules is depicted in
Figure~\ref{fig:hardware-definition-graph}.

\begin{figure}[h]
\centering
\includegraphics[width=.7\textwidth]{diagramaEstrutural.jpg}
\caption{Dependencies between modules of $\mathit{HARDWARE\_LIBRARY}$ and
platform model.}
\label{fig:hardware-definition-graph}

\end{figure}

Now, We will explain each of these modules and present them in detail.

\subsection{Definitions to represent and manipulate bits}

@@

First, bits are modeled as the set of integers $\{0, 1\}$:\\

$
\begin{array}{l}
\mathit{BIT} = \mathit{0..1} \\ 
\end{array}
$

The negation is a unary function on bits and it is defined as:

$
\begin{array}{l}
\mathit{bit\_not}  \in  \mathit{BIT}  \fun  \mathit{BIT}  \land \\
\forall ( \mathit{bb}). (\mathit{bb} \in \mathit{BIT} \implies \mathit{bit\_not}(\mathit{bb}) =
1-\mathit{bb})\\
\end{array}
$

@@

$
\begin{array}{l}
 \mathit{bit\_not}(0) = 1;  \mathit{bit\_not}(1) = 0; \\
\forall (\mathit{bb}).(\mathit{bb} \in \mathit{BIT} \implies \mathit{bit\_not}(\mathit{bit\_not}(\mathit{bb})) = \mathit{bb});
\end{array}
$

Conjunction is a unary function on bits and it is defined as:

@@

$
\begin{array}{l}
 \mathit{bit\_and}(0,0) = 0;  \mathit{bit\_and}(0,1) = 0; \\
 \mathit{bit\_and}(1,0) = 0;  \mathit{bit\_and}(1,1) = 1; \\
\forall (\mathit{b1},\mathit{b2}).(\mathit{b1} \in \mathit{BIT} \land \mathit{b2} \in \mathit{BIT} \implies \\
\quad (\mathit{bit\_and}(\mathit{b1}, \mathit{b2}) = \mathit{bit\_and}(\mathit{b2},\mathit{b1}))); \\
\forall (\mathit{b1},\mathit{b2},\mathit{b3}).(\mathit{b1} \in \mathit{BIT} \land  \mathit{b2} \in \mathit{BIT} \land \mathit{b3} \in \mathit{BIT} \implies \\
\quad (\mathit{bit\_and}(\mathit{b1}, \mathit{bit\_and}(\mathit{b2},\mathit{b3})) = \mathit{bit\_and}(\mathit{bit\_and}(\mathit{b1},\mathit{b2}),\mathit{b3}))); \\
\forall (\mathit{b1}).(\mathit{b1} \in \mathit{BIT} \implies (\mathit{bit\_and}(\mathit{b1}, 1) = \mathit{b1})); \\
\forall (\mathit{b1}).(\mathit{b1} \in \mathit{BIT} \implies (\mathit{bit\_and}(\mathit{b1}, 0) = 0));
\end{array}
$

@@

\subsection{Representation and manipulation of bit vectors}

Sequences are pre-defined in B, as functions whose the domain is an integer range with lower bound 1 (one). Indices
in bit vectors usually range from 0 (zero) upwards and the model we propose obeys this convention by making an
one-position shift where necessary.  We thus define bit vectors as non-empty sequences of bits, and
$\mathit{BIT\_VECTOR}$ is the set of all such sequences:

$
\begin{array}{l}
\mathit{BIT\_VECTOR} = \seq (\mathit{BIT})
\end{array}
$

@@

$
\begin{array}{l}
\mathit{bv\_size} \in \mathit{BIT\_VECTOR} \fun \nat_1 \land \\
\mathit{bv\_size} = \lambda bv \bullet (bv \in \mathit{BIT\_VECTOR} \mid \mathbf{size}(bv))
\end{array}
$

@@

$
\begin{array}{l}
\mathit{bv\_set} \in \mathit{BIT\_VECTOR} \times \nat \fun \mathit{BIT\_VECTOR} \land \\
\mathit{bv\_set} = \lambda v, n \bullet (v \in \mathit{BIT\_VECTOR} \land 0 \in \nat \land n < \mathit{bv\_size}(v) \mid v \ovr \{ n+1 \mapsto 1 \})
\end{array}
$

@@

\hspace*{0.00in} \it bv\_catenate  $\in$  \it BIT\_VECTOR  $\times$  \it BIT\_VECTOR  $\fun$ \it BIT\_VECTOR $\land$

\hspace*{0.00in} \it bv\_catenate \rm =  $\lambda$  v\rm ,\it w \rm . \rm (\it v  $\in$  \it BIT\_VECTOR $\land$ \it w $\in$  \it BIT\_VECTOR  $\mid$  \it v $\cat$ \it w\rm )

@@

$
\begin{array}{l}
\mathit{bv\_not} \in \mathit{BIT\_VECTOR} \fun \mathit{BIT\_VECTOR} \land \\
\mathit{bv\_not} = \lambda v \bullet (v \in \mathit{BIT\_VECTOR} \mid \\
\quad \lambda i \bullet 0 .. \mathit{bv\_size}(v)-1 \mid \mathit{bit\_not}(v(i))) \land \\
\mathit{bv\_and} \in \mathit{BIT\_VECTOR} \times \mathit{BIT\_VECTOR} \fun \mathit{BIT\_VECTOR} \land \\
\mathit{bv\_and} = \lambda v_1, v_2 \bullet (v_1 \in \mathit{BIT\_VECTOR} \land v_2 \in \mathit{BIT\_VECTOR} \land \\
\quad \mathit{bv\_size}(v_1) = \mathit{bv\_size}(v_2) \mid \lambda i \bullet 0 .. \mathit{bv\_size}(v_1)-1 \mid \mathit{bit\_and}(v_1(i), v_2(i)))
\end{array}
$

We provide several lemmas on bit vector operations. These lemmas
express properties on the size of the result of the operations
as well as classical algebraic properties such as associativity
and commutativity. For example:

$
\begin{array}{l}
\forall v \bullet (v \in \mathit{BIT\_VECTOR} \implies \mathit{bv\_size}(\mathit{bv\_not}(v)) = \mathit{bv\_size}(v)) \\

\forall v \bullet (v \in \mathit{BIT\_VECTOR} \implies \mathit{bv\_not}(\mathit{bv\_not}(v)) = v) \\

\forall v_1, v_2 \bullet (\{v_1, v_2\} \subseteq \mathit{BIT\_VECTOR} \land \mathit{bv\_size}(v_1) = \mathit{bv\_size}(v_2) \implies \\
\quad \mathit{bv\_size}(\mathit{bv\_and}(v_1, v_2)) = \mathit{bv\_size}(v_1)) \\

\forall v_1, v_2 \bullet (\{v_1, v_2\} \subseteq \mathit{BIT\_VECTOR} \land \mathit{bv\_size}(v_1) = \mathit{bv\_size}(v_2) \implies \\
\quad \mathit{bv\_size}(\mathit{bv\_and}(v_1, v_2)) = \mathit{bv\_size}(v_2)) \\

\forall v_1, v_2 \bullet (\{v_1, v_2\} \subseteq \mathit{BIT\_VECTOR} \land \mathit{bv\_size}(v_1) = \mathit{bv\_size}(v_2) \implies \\
\quad \mathit{bv\_size}(\mathit{bv\_and}(v_1, v_2)) = \mathit{bv\_and}(v_2, v_1)) \\

\forall v_1, v_2, v_3 \bullet (\{v_1, v_2, v_3\} \subseteq \mathit{BIT\_VECTOR}
\land \mathit{bv\_size}(v_1) = \mathit{bv\_size}(v_2) \land \\
\quad \mathit{bv\_size}(v_1) = \mathit{bv\_size}(v_3) \implies \\ 
\quad \mathit{bv\_and}(\mathit{bv\_and}(v_1, v_2), v_3) = \mathit{bv\_and}(v_1,
\mathit{bv\_and}(v_2, v_3))) \\

\forall v \bullet (v \in \mathit{BIT\_VECTOR} \implies \\
\quad \mathit{bv\_and}(v, \mathit{bv\_zero}(\mathit{bv\_size}(v))) = \mathit{bv\_zero}(\mathit{bv\_size}(v))) \\

\forall v \bullet (v \in \mathit{BIT\_VECTOR} \implies \\
\quad \mathit{bv\_and}(v, \mathit{bv\_one}(\mathit{bv\_size}(v))) = v) \\

\end{array}
$

\subsection{Modelling bytes and bit vector of length 16}[I need review the english]

@@

% $
% \begin{array}{l}
% \mathit{BYTE\_WIDTH} = 8 \land \mathit{BYTE\_INDEX} = 0..(\mathit{BYTE\_WIDTH}-1) \land \\
% \mathit{BYTE} \subseteq \mathit{BIT\_VECTOR} \land \mathit{BYTE} = \{ v \in \mathit{BIT\_VECTOR} \land \mathit{bv\_size}(v) = \mathit{BYTE\_WIDTH}\} \land \\
% \mathit{BYTE\_ZERO} \in \mathit{BYTE} \land \mathit{BYTE} = \mathit{BYTE\_INDEX} \times \{0\}
% \end{array}
% $
% Version 2

\hspace*{0.0in}\it BYTE\_WIDTH \rm = 8 $\land$ \it BYTE\_INDEX \rm = 1 $\upto$ \rm (\it BYTE\_WIDTH\rm )$\land$

@@

\hspace*{0.0in}\it BYTE\_ZERO  $\in$  \it BYTE  $\land$

\hspace*{0.0in}\it BYTE\_ZERO \rm = \it BYTE\_INDEX  $\times$  \rm \{\rm 0\rm \}

The $\mathit{BYTE\_INDEX}$ is domain of byte modeled. it start at 1 to  obey a definition of sequences from B
method. However, the functions from byte encapsulate the acces and these functions use the
$\mathit{PHYS\_BYTE\_INDEX}$. The $\mathit{BYTE}$ type is a specializated type from $\mathit{BIT\_VECTOR}$, but it
has a size limit. Some other more specific definitions are created to help the developer. For example the constant
$\mathit{BYTE\_ZERO}$ and some below functions.

\begin{itemize}
  \item \it is\_zero $\in$ BYTE  $\fun$   BIT \rm - It returns 1 if the received byte is zero, otherwise
  returns 0.

  \item \it rotateright  $\in$  \it BYTE  $\fun$  \it BYTE \rm - It receives a byte and returns its rotated to
  right. There is a similar function named $\mathit{rotateleft}$.

By the very similar way, the type $\mathit{BV16}$ is created.

\subsection{Bit vector arithmetics}

Bit vectors are used to represent and combine numbers: integer ranges
(signed or unsigned) and floating point numbers.

Our library defines a function $\mathit{bv\_to\_nat}$ that maps bit
vectors to natural numbers:

$
\begin{array}{l}
\mathit{bv\_to\_nat} \in \mathit{BIT\_VECTOR} \fun \nat \land \\
\mathit{bv\_to\_nat} = \lambda v \bullet (v \in \mathit{BIT\_VECTOR} \mid \sum i \bullet (i \in \dom(v) \bullet v(i) \times 2^i))
\end{array}
$

We provide the following lemma:

$
\begin{array}{l}
\forall n \bullet (n \in \nat_1 \implies \mathit{bv\_to\_nat}(\mathit{bv\_zero}(n)) = 0)
\end{array}
$

\subsection{Basics data types}

There are data types commonly used in the microcontrollers. They are placed in
the project Types Librarys and define functions to manipulate data on levels of
basic types from hardware. However, in general is needed to detail or redefines
the types and its functions for each platform. Then, the specific concepts are
defined in the project's own platform.

There are six common basics data types, see details in table below.

\begin{tabular}{|p{3.5cm}|c|c|c|}\hline
 $Type\ Name$ & $Range$ & $Physical\ Size $\\ 
\hline
 UCHAR& 0..255 & 1 byte\\ \hline SCHAR& -128..127 & 1 byte\\ \hline BYTE & -- & 1
 byte\\ \hline USHORTINT & 0..65.535 & 2 byte\\ \hline SSHORTINT &
 -32.768..32.767 & 2 byte\\ \hline BV16 & -- & 2 byte\\ \hline
\end{tabular}

There is, for example, the function $\mathit{bv\_to\_nat}$ that is specialized
to $\mathit{byte\_uchar}$. How the set $\mathit{BYTE}$ is a subset of the
$\mathit{BIT\_VECTOR}$, then this function can defined as follows:

$
\begin{array}{l}
\mathit{byte\_uchar} = \lambda (v) \bullet ( v \in BYTE | bv\_nat(v) )
\end{array}
$

Usually, the definitions of the functions of the module $\mathit{TYPES}$ are the
re-used from the basic functions of the hardware library. This provides greater
confidence and facilitates the process of proof, because the prover can use the
defined previous lemma.

A more specific model of $\mathit{byte\_uchar}$ functions is:

$
\begin{array}{l}
\mathit{byte\_uchar} \in \mathit{BYTE}  \fun  \mathit{UCHAR}  \land \\
   \mathit{byte\_uchar} =  \lambda ( v0 )  \bullet  
   ( v0 \in \mathit{BYTE} |  2^{7}* bv\_get( v0 ,7 ) + 2^{6} * bv\_get( v0 ,6 ) + 2^{5}* bv\_get( v0 , 5   )\\ 
   + 2^{4}* bv\_get( v0 , 4 ) + 2^{3}*bv\_get( v0 , 3 ) + 2^{2}* bv\_get( v0 , 2 ) + 2 * bv\_get( v0 , 1 ) +  bv\_get( v0 ,0) )

\end{array}
$

The inverse function is easily defined as $\mathit{uchar\_byte}$.

$
\begin{array}{l}
\mathit{uchar\_byte} \in \mathit{UCHAR}  \fun  \mathit{BYTE}  \land \\
   \mathit{uchar\_byte} = \  ^{-1}(\mathit{byte\_uchar}
   )
\end{array}
$

We also created the following lemmas:

$
\begin{array}{l}
 \forall (val) \bullet (val \in \mathit{UCHAR} |
 \mathit{byte\_uchar}(\mathit{uchar\_byte}(val)) = val) \land\\
 \forall (by) \bullet (by \in \mathit{BYTE} |
 \mathit{uchar\_byte}(\mathit{byte\_uchar}(by)) = by)
\end{array}
$

@@

\section{Description of the B model from Z80}
\label{sec:z80}

The \textit{Z80} is an important microcontroller developed in 1976 by \textit{Zilog}. In this time, It got a great
acceptance because suported all set instruction of \textit{8080}, one of the predecessors of the \textit{Pentium}
from \textit{Intel Corporation}. Then, the Z80 has the set instructions similar to the x86 architecture, it helps 
to build the model of x86 and it provides good features.

The main module includes a instance of memory module and accesses the definitions from: basic data types modules,
\textit{ALU} and \textit{Power2} \footnote{The \textit{Power2} module has the basics definitions to help the
theorems prover about calculus of power.}.

@@

\hspace*{0.10in}\it SSHORT\_DEFINITION ,\it USHORT\_DEFINITION, \it POWER2
\end{sloppypar}

>>>>

\documentclass{llncs}
%\documentclass{article}

\usepackage{graphicx,array}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{epsfig}
\usepackage{float}
%\usepackage{supertabular} % used in B Symbols

@@

\title{Formal Construction of Microcontroller Instruction Set Model Using B}

\hyphenation{se-pa-ra-te}

\author{ David D\'{e}harbe\inst{1}, Val\'{e}rio Medeiros Jr\inst{2} }
\institute{Federal University of Rio Grande do Norte, Natal RN 59078-970, Brazil\inst{1} and\\  
Federal Institute of Education, Science and Technology of Rio Grande do Norte, Natal RN
59015-000, Brazil\inst{2}}

@@

\begin{abstract}
  [To redo]
  This paper describes an approach to model the functional aspects of
  the instruction set of microcontroller platforms using the notation
  of the B method. The paper presents specifically the case of the Z80 
  platform. This work is a contribution towards the extension of the
  B method to handle developments up to assembly level code.
\end{abstract}
%
%\keywords{Fomal Methods, microcontroller Verification, Hardware}

\section{Introduction}
[To update]
The B method~\cite{Abrial} supports the construction of safety systems
models by verification of proofs that guarantees its correctness. So,
an initial abstract model of the system requirements is defined and
then it is refined until the implementation model. Development
environments based on the B method also include source code generators
for programming languages, but the result of this translation
cannot be compared by formal means. The paper \cite{Dantas_SBMF08} presented
recently an approach to extend the scope of the B method up to the
assembly level language. One key component of this approach
is to build, within the framework of the B method, formal models of
the instruction set of such assembly languages.

This work gives an overview of the formal modelling of the instruction
set of the Z80 microcontroller~\cite{Z80_manual}\footnote{The
  interested reader in more details is invited to visit our repository
  at: http://code.google.com/p/b2asm.}. Using the responsibility
division mechanism provided by B, auxiliary libraries of basic modules
were developed as part of the construction of microcontroller
model. Such library has many definitions about common concepts used in
the microcontrollers; besides the Z80 model, it is used by two other
microcontrollers models that are under way.

Other possible uses of a formal model of a microcontroller
instruction set include documentation, the construction of simulators,
and be possibly the starting point of a verification effort for the
actual implementation of a Z80 design. Moreover the model of the
instruction set could be instrumented with non-functional aspects,
such as the number of cycles it takes to execute an instruction, to
prove lower and upper bounds on the execution time of a routine. The
goal of this project, though, is to provide a basis for the generation
of software artifacts at the assembly level that are amenable to 
refinement verification within the B method.

This paper is focused on the presentation of the Z80 model, including
elementary libraries to describe hardware aspects. The paper is
structured as follows. Section~\ref{sec:B_method} provides a short
introduction to the B method. Section~\ref{sec:models} presents the
elementary libraries and the modelling of some elements common to
microcontrollers.  Section~\ref{sec:z80} presents the B model of the
Z80 instruction set. Section ~\ref{sec:Proofs} provides some
information on the proof effort needed to analyze the presented
models. Related work is discussed in
Section~\ref{sec:relatedworks}. Finally, the last section is devoted
to the conclusions.

\section{Introduction to the B Method}
\label{sec:B_method}

The B method for software development~\cite{Abrial} is based on the B Abstract
Machine Notation (AMN) and the use of formally proved refinements up to a
specification sufficiently concrete that programming code can be automatically
generated from it. Its mathematical basis consists of first order logic, integer
arithmetic and set theory, and its corresponding constructs are similar to those
of the Z notation.

A B specification is structured in modules. A module defines a set of valid
states, including a set of initial states, and operations that may provoke a
transition between states. The design process starts with a module with a
so-called functional model of the system under development. In this initial
modelling stage, the B method requires that the user proves that, in a machine,
all the its initial states are valid, and that operations do not define
transitions from valid states to invalid states.

Essentially, a B module contains two main parts: a header and the available
operations. Figure~\ref{fig:maqB} has a very basic example. The clause
$\mathit{MACHINE}$ has the name of module.  The next two clauses respectively
reference external modules and create an instance of an external module. The
$\mathit{VARIABLES}$ clauses declares the name of the variables that compose the
state of the machine. Next, the $\mathit{INVARIANT}$ clause defines the type and
other restrictions on the variables. The $\mathit{INITIALIZATION} $ specifies
the initial states. Finally, operations correspond to the transitions between
states of the machine.

\begin{figure}
  % \begin{small}
  $$
  \begin{array}{lcl}
    \begin{array}[t]{l}
      \MACHINE \quad  \mathit{micro} \\
      \SEES \quad \mathit{TYPES}, \mathit{ALU} \\
      \INCLUDES  \quad \mathit{MEMORY} \\
      \VARIABLES \quad    \mathit{pc}  \\
      \INVARIANT \quad \mathit{pc} \in \mathit{INSTRUCTION} 
    \end{array}
    & \hspace*{1cm} &
    \begin{array}[t]{l}
      \INITIALISATION  \mathit{pc} := 0 \\
      \OPERATIONS\\
      \mathit{JMP} \mathit{( jump )} = \\
      \quad \PRE \mathit{jump} \in \mathit{INSTRUCTION}\\ 
      \quad \THEN \mathit{pc} := \mathit{jump}\\  
      \quad \END \\
      \END
    \end{array}
  \end{array}
  $$
\caption{A very basic B machine.}
\label{fig:maqB}
\end{figure}

\section{Model structure and basic components}
\label{sec:models}

We have been developed a reusable set of basic definitions to model hardware concepts and data types
concepts. These definitions are grouped into two separate development projects and are available as
libraries. A third project is devoted to the higher-level aspects of the platform. Thus, the workspace is
composed of: a power library\footnote{The power library has the basics definitions to help the theorems
prover about simple calculus of power, especially power of two. It has constants, that represents all
results of power function needed, and the definition of power function.}, a hardware library, a types
library and a project for the specific platform, in this case the Z80. The corresponding dependency
diagram is depicted in Figure~\ref{fig:hardware-definition-graph}; information specific to each project is
presented in the following.

\begin{figure}[h] \centering
\includegraphics[width=.92\textwidth]{diagramaEstrutural_vertical.png}
 \caption{Dependency diagram of the Z80 model.}
\label{fig:hardware-definition-graph}
\end{figure}

\subsection{Bit Representation and Manipulation}

@@

First, bits are modelled as a set of integers: $\mathit{BIT} =
\mathit{0..1}$. The negation is an unary function on bits and it is
defined as:

$
\begin{array}{l}
\mathit{bit\_not}  \in  \mathit{BIT}  \fun  \mathit{BIT}  \land \forall ( \mathit{bb}). (\mathit{bb} \in \mathit{BIT} \implies \mathit{bit\_not}(\mathit{bb}) =
1-\mathit{bb})\\
\end{array}
$

@@

$
\begin{array}{l}
%  \mathit{bit\_not}(0) = 1;  \mathit{bit\_not}(1) = 0; \\
\forall (\mathit{bb}).(\mathit{bb} \in \mathit{BIT} \implies \mathit{bit\_not}(\mathit{bit\_not}(\mathit{bb})) = \mathit{bb})
\end{array}
$

Conjunction is an unary function on bits and it is defined as:

@@

$
\begin{array}{l}
%  \mathit{bit\_and}(0,0) = 0;  \mathit{bit\_and}(0,1) = 0; \\
%  \mathit{bit\_and}(1,0) = 0;  \mathit{bit\_and}(1,1) = 1; \\
\forall (\mathit{b1},\mathit{b2}).(\mathit{b1} \in \mathit{BIT} \land \mathit{b2} \in \mathit{BIT} \implies \\
\quad (\mathit{bit\_and}(\mathit{b1}, \mathit{b2}) = \mathit{bit\_and}(\mathit{b2},\mathit{b1})))\land \\
\forall (\mathit{b1},\mathit{b2},\mathit{b3}).(\mathit{b1} \in \mathit{BIT} \land  \mathit{b2} \in \mathit{BIT} \land \mathit{b3} \in \mathit{BIT} \implies \\
\quad (\mathit{bit\_and}(\mathit{b1}, \mathit{bit\_and}(\mathit{b2},\mathit{b3})) = \mathit{bit\_and}(\mathit{bit\_and}(\mathit{b1},\mathit{b2}),\mathit{b3})))\\
% \forall (\mathit{b1}).(\mathit{b1} \in \mathit{BIT} \implies (\mathit{bit\_and}(\mathit{b1}, 1) = \mathit{b1})); \\
% \forall (\mathit{b1}).(\mathit{b1} \in \mathit{BIT} \implies (\mathit{bit\_and}(\mathit{b1}, 0) = 0));
\end{array}
$

@@

\subsection{Representation and Manipulation of Bit Vectors}

Sequences are pre-defined in B, as functions whose the domain is an
integer range with lower bound 1 (one). Indices in bit vectors usually
range from 0 (zero) upwards and the model we propose obeys this
convention by making an one-position shift where necessary. This shift
is important to use the predefined functions of sequences. We thus
define bit vectors as non-empty sequences of bits, and
$\mathit{BIT\_VECTOR}$ is the set of all such sequences:
$\mathit{BIT\_VECTOR} = \seq1(\mathit{BIT})$.

@@

$
\begin{array}{l}
\mathit{bv\_size} \in \mathit{BIT\_VECTOR} \fun \nat_1 \land \\
\mathit{bv\_size} = \lambda bv . (bv \in \mathit{BIT\_VECTOR} \mid \mathbf{size}(bv))
\end{array}
$

@@

$
\begin{array}{l}
\mathit{bv\_set} \in \mathit{BIT\_VECTOR} \times \nat \fun \mathit{BIT\_VECTOR} \land \mathit{bv\_set} =\\
\lambda v, n . (v \in \mathit{BIT\_VECTOR} \land n \in \nat \land n <\mathit{bv\_size}(v)
\mid v \lover \{ n+1 \mapsto 1 \})
\end{array}
$

@@

\hspace*{0.00in} \it bv\_catenate  $\in$  \it BIT\_VECTOR  $\times$  \it BIT\_VECTOR  $\fun$ \it
BIT\_VECTOR $\land$\\
\hspace*{0.00in} \it bv\_catenate \rm =  $\lambda$  v\rm ,\it w \rm . \rm (\it v 
$\in$  \it BIT\_VECTOR $\land$ \it w $\in$  \it BIT\_VECTOR  $\mid$  \it v\^ \it w\rm ) %$\cat$

@@

$
\begin{array}{l}
\mathit{bv\_not} \in \mathit{BIT\_VECTOR} \fun \mathit{BIT\_VECTOR} \land \\
\mathit{bv\_not} = \lambda v . (v \in \mathit{BIT\_VECTOR} \mid \quad \lambda i . (1 .. \mathit{bv\_size}(v)) \mid \mathit{bit\_not}(v(i))) \land \\
\mathit{bv\_and} \in \mathit{BIT\_VECTOR} \times \mathit{BIT\_VECTOR} \fun \mathit{BIT\_VECTOR} \land \\
\mathit{bv\_and} = \lambda v_1, v_2 . (v_1 \in \mathit{BIT\_VECTOR} \land v_2 \in \mathit{BIT\_VECTOR} \land \\
\quad \mathit{bv\_size}(v_1) = \mathit{bv\_size}(v_2) \mid \lambda i . (1 .. \mathit{bv\_size}(v_1)) \mid
\mathit{bit\_and}(v_1(i), v_2(i)))
\end{array}
$

We provide several lemmas on bit vector operations. These lemmas
express properties on the size of the result of the operations
as well as classical algebraic properties such as associativity
and commutativity.

\subsection{Modelling Bytes and Bit Vectors of Length 16}

@@

\hspace*{0.0in}\it BYTE\_WIDTH \rm = 8 $\land$ \it BYTE\_INDEX \rm = 1 $\upto$ \rm  BYTE\_WIDTH\rm $\land$

@@

\hspace*{0.0in}\it BYTE\_ZERO  $\in$  \it BYTE  $\land$ \it BYTE\_ZERO \rm = \it BYTE\_INDEX  $\times$  \rm \{\rm 0\rm \}

The $\mathit{BYTE\_INDEX}$ is the domain of the functions modelling bytes. It starts at 1 to obey a
definition of sequences from B. However, it is common in hardware architectures to start indexing from
zero. The definition $\mathit{PHYS\_BYTE\_INDEX}$ is used to provide functionalities obeying this
convention. The $\mathit{BYTE}$ type is a specialized type from $\mathit{BIT\_VECTOR}$, but it has a size
limit. Other specific definitions are provided to facilitate further modelling: the type $\mathit{BV16}$
is created for bit vector of length 16 in a similar way.

\subsection{Bit Vector Arithmetics}

Bit vectors are used to represent and combine numbers: integer ranges (signed or unsigned). Therefore, our
library includes functions to manipulate such data, for example, the function $\mathit{bv\_to\_nat}$ that
maps bit vectors to natural numbers:

$
\begin{array}{l}
\mathit{bv\_to\_nat} \in \mathit{BIT\_VECTOR} \fun \nat \land \\
\mathit{bv\_to\_nat} = \lambda v . (v \in \mathit{BIT\_VECTOR} \mid \sum i . (i \in \dom(v) . v(i)
\times 2^{i-1}))
\end{array}
$

An  associated lemma is: $\forall n . (n \in \nat_1 \implies \mathit{bv\_to\_nat}(\mathit{nat\_to\_bv}(n)) = n)$

\subsection{Basics Data Types}

The instruction set of microcontrollers usually have common data types. These types are placed in the
types library. Each type module has functions to manipulate and convert its data. There are six common
basics data types represented by modules, see details in table~\ref{tab:types}.

\begin{table}
\caption{Descriptions of basic data types}
\label{tab:types}

\begin{center}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
 $Type\ Name$ & UCHAR & SCHAR & USHORTINT & SSHORTINT  & BYTE & BV16 \\\hline
 $Range$ & 0..255 & -128..127 & 0..65.535  & -32.768..32.767  & -- & --\\ \hline
 $Physical\ Size $ & 1 byte & 1 byte & 2 bytes & 2 bytes &  1 bytes & 2 bytes \\ \hline
\end{tabular}
\end{center}
\end{table}

Usually, each type module just needs to instantiate concepts that were already defined in the hardware
modelling library.  For example, the function $\mathit{bv\_to\_nat}$ from bit vector arithmetics is
specialized to $\mathit{byte\_uchar}$. As the set $\mathit{BYTE}$ is a subset of the
$\mathit{BIT\_VECTOR}$, this function can defined as follows:

$
\begin{array}{l}
\mathit{byte\_uchar} \in \mathit{BYTE} \fun \nat \land \\
\mathit{byte\_uchar} = \lambda (v) . ( v \in BYTE | bv\_to\_nat(v) )
\end{array}
$

The definitions of the library types reuse the basic definitions from the hardware library. This provides
greater confidence and facilitates the proof process, because the prover can reuse the previously defined
lemma.

The inverse function $\mathit{uchar\_byte}$ is easily defined:

$
\begin{array}{l}
\mathit{uchar\_byte} \in \mathit{UCHAR}  \fun  \mathit{BYTE}  \land \\
   \mathit{uchar\_byte} = \ (\mathit{byte\_uchar}) ^{-1}
\end{array}
$

% We also created the following lemmas:
%
% $
% \begin{array}{l}
%  \forall (val) . (val \in \mathit{UCHAR} |
%  \mathit{byte\_uchar}(\mathit{uchar\_byte}(val)) = val) \land\\
%  \forall (by) . (by \in \mathit{BYTE} |
%  \mathit{uchar\_byte}(\mathit{byte\_uchar}(by)) = by)
% \end{array}
% $

@@

\section{Description of the Z80 B model }
\label{sec:z80}

The \textit{Z80} is a CISC microcontroller developed by \textit{Zilog}~\cite{Z80_manual}. It supports 158
different instructions, including all the 78 from Intel 8080 microprocessor, and all of them were
specified. These instructions are classified into these categories: load and exchange; block transfer and
search; arithmetic and logical; rotate and shift; bit manipulation; jump, call and return; input/output;
and basic cpu control.

The main module includes an instance of the memory module and accesses the definitions from basic data
types modules and the \textit{ALU} module.

@@

\hspace*{0.10in}\it SSHORT\_DEFINITION ,\it USHORT\_DEFINITION
\end{sloppypar}

% Chacteristics of microcontroller
 The internal registers contain 176 bits of reading/writing memory that are represented by
identifiers used as parameters in the instructions . It includes two sets of six general purpose
registers which may be used individually as 8-bits registers or as 16-bits register pairs.  The working registers
are represented by variable $\mathit{rgs8}$. The domain of $\mathit{rgs8}$ ($\mathit{id\_regs8}$) is a set
formed by identifiers of registers of 8 bits. These registers can be accessed in pairs, forming 16-bits,
resulting in another set of identifiers of 16-bits registers, named $\mathit{id\_reg16}$.
